Fast and scalable Gaussian process modeling of stellar variability \vspace{-3\baselineskip}

### 2016-10-25 paper/ms.tex
Added kernel approximation plot; cleaned up notation.

The functional form of $K_j$ may be used to approximate some commonly used Kernel components: radial
kernels can be approximated with $\beta_I=0$, while periodic kernels can be approximated by
$\beta_I \ne 0$.  The squared exponential (Gaussian) kernel and the Mat\'ern kernels are commonly
used radial kernels, while the cosine kernel and the exponential-sine-squared kernels are
commonly used periodic kernels.  Linear combinations of our basis of damped cosine kernels may be 
used to approximate each of these kernels, with various degrees of success:
\begin{itemize}
\item {\bf Constant kernel} A constant kernel component can be used to capture long-timescale
correlations in data, and can be represented with $\beta =  0$.
\item {\bf Exponential kernel.} This is exactly
represented with $\beta_I=0$.
\item {\bf Cosine kernel.} The cosine kernel may be represented exactly with $\beta_R=0$.
\item {\bf Squared exponential.} The squared exponential or Gaussian kernel, $G(z)=e^{-z^2/2}$,
has the behavior of declining steeply at large time separation, is an even function, and has $c_1$ = 0.
Nevertheless, it may be well approximated by the sum of three exponential kernels:
\begin{eqnarray}
G(z) &\approx& 2.751807 e^{-1.531250 z} \cos{(0.341489 z)} \\
 &-& 0.354626 e^{-4.753504 z} \cos{(2.042572 z)}\\
 &-& 1.396834 e^{-1.796743 z} \cos{(1.764168 z)},
\end{eqnarray}
which has an accuracy of better than $<4\times 10^{-4}$ for all z.
%[2.751807496353032,1.5312499408182438,-0.34148888180984543,-0.35462563655835033,4.753504292338051,2.0425716395901157,-1.3968344706423566,1.7967431129750422,-1.7641684073374964
\item {\bf Mat\'ern $p+1/2$ kernel.}  The Mat\'ern kernel is given by:
\begin{equation}
C_{p+1/2}(x) = \sigma^2 e^{-\sqrt{(2p+1)}x} \frac{\Gamma(p+1)}{\Gamma(2p+1)}
\sum_{i=0}^p \frac{(p+i)!}{i!(p-i)!} (\sqrt{(8p+4)}x)^{p-i},
\end{equation} 
where $p$ is a positive integer.  In the limit $p\rightarrow \infty$, this becomes the
squared exponential kernel: $C_\infty(x) = \sigma^2 e^{-x^2/2}$.  But, for small
values of $p$, the Mat\'ern kernel may be approximated well by the sum of
$2p$ exponential kernels.  The exponential kernel is the Mat\'ern 1/2 kernel, while
$p=1$ is relevant to stellar variability, which we describe next.
\item {\bf Mat\'ern 3/2 kernel.}  The Mat\'ern $3/2$ kernel is given by: 
\begin{equation}
K_{3/2}(\tau,\tau_0) = (1+x)e^{-x},
\end{equation} 
where $x = \sqrt{3}\vert \tau/\tau_0\vert$.  This kernel has a Fourier Transform\footnote{Using the normalization convention: 
$\hat f(\omega) = \frac{1}{\sqrt{2\pi}} \int_{-\infty}^\infty f(t) e^{i\omega t} dt$}  of
\begin{equation}
\hat K_{3/2}(\omega) = \frac{\tau_0 6^{3/2}}{\pi^{1/2}} (3+\tau_0^2\omega^2)^{-2},
\end{equation}
where $\omega = 2\pi f$.  This kernel may be well approximated by the sum of two exponential kernels: 
\begin{equation}
K_{3/2}(\tau,\tau_0) \approx (1-b)e^{-x}+be^{-x\frac{b-1}{b}},
\end{equation} 
where $b \gg 1$ is a dimensionless parameter; this approximation is exact in the limit $b \rightarrow \infty$.  For
large values of $b$, the maximum error on this scales as $2/(e^2b) \approx 0.27/b$; for
$b=10$, the error in this approximation is $<3$\%.  Note that the two exponential terms
have opposite signs, which may lead to numerical instability for values of $b$ that
are large.

One advantage of the Mat\'ern 3/2 kernel is that it is smooth across
$\tau=0$ since its derivative is zero at $\tau=0$.  The exponential kernel {\it does not}
have this property, which leads to qualitatively different noise properties on small
timescales due to the sharpness of the exponential kernel at $\tau=0$.  The Mat\'ern 3/2 kernel
turns out to be similar to the observed properties of stellar activity and granulation, and thus the approximate
formula is simultaneously physically relevant and computationally convenient.  
The power spectrum of a single exponential scales as $f^{-2}$ at high frequency,
while the Mat\'ern kernel scales as $f^{-4}$ which more closely matches the power
spectrum due to activity and granulation in the Sun \citep{2004A&A...414.1139A,2009A&A...495..979M}.
\item {\bf Mat\'ern 5/2 kernel.}  The Mat\'ern 5/2 kernel is given by
\begin{equation}
C_{5/2}(y) = \sigma^2 (1+y+y^2/3)e^{-y},
\end{equation}
where $y = \sqrt{5}x$.  The Fourier transform is:
\begin{equation}
\hat K_{5/2}(\omega) = \frac{2\tau_0 10^{5/2}}{3\pi^{1/2}} (5+\tau_0^2\omega^2)^{-3}.
\end{equation}
This kernel may be approximated well by the sum of four exponential kernels
\begin{eqnarray}
K_{5/2}(y) &\approx& -147.6965573 e^{-1.0097610 y} + 85.8875034 e^{-1.0621734 y}\\
 &-&4.6540017 e^{-0.8353745 y} + 67.4630744 e^{-0.9160394 y},
\end{eqnarray}
which is accurate to $<4\times 10^{-6}$ for $y \le 10$.
The Fourier transform has a steeper dependence at large frequencies, $f^{-6}$, and as we are not aware of a stellar
power spectrum that requires a steeper spectrum than this, we forgo approximating further Mat\'ern kernels.
\item {\bf Exponential-sine-square kernel.}  This kernel is a periodic, positive valued kernel
which has been used to model quasi-periodic stellar variability due to star spots.   It is
given by $E(w,\Gamma) = e^{-\Gamma \sin^2{w}}$, which may be expanded as a series in $\cos{(2kw)}$,
with $k \in \mathbb{N}$.  We have expanded this to $k=4$, and fit for the coefficients, giving:  
\begin{equation}
E(w,\Gamma) \approx \sum_{k=0}^4 \left(c_{k,1}+c_{k,2} e^{-c_{k,4}\Gamma^{c_{k,3}}}\right) \cos{(2kw)},
\end{equation}
with $c_{0,1}=0.24999, c_{0,2}=0.752803, c_{0,3}=0.960477, c_{0,4}=0.643393$ for $k=0$,
$c_{1,1}=0.450851, c_{1,2}=-0.445338, c_{1,3}=0.960477, c_{1,4}= 1.18102$ for $k=1$,
$c_{2,1}=0.206593, c_{2,2}=-0.207361, c_{2,3}=1.65902, c_{2,4}= 0.210641$ for $k=2$,
$c_{3,1}=0.0728449, c_{3,2}=-0.0729381, c_{3,3}=2.43149, c_{3,4}= 0.0471224$ for $k=3$, and
$c_{4,1}=0.019204, c_{4,2}=-0.0192129, c_{4,3}=3.27977, c_{4,4}= 0.0114198$ for $k=4$.
This approximation is accurate to better than 0.5\% for $\Gamma < 3$.
\end{itemize}

{\color{red} Should we include some plots of these kernels, with the approximations?}

>>>>

The functional form of $K_j$ may be used to approximate some commonly used Kernel components: radial
kernels can be approximated with $\beta_I=0$, while periodic kernels can be approximated by
$\beta_I \ne 0$.  The squared exponential (Gaussian) kernel and the Mat\'ern kernels are commonly
used radial kernels, while the cosine kernel and the exponential-sine-squared kernels are
commonly used periodic kernels.  Linear combinations of our basis of damped cosine kernels may be 
used to approximate each of these kernels, with various degrees of success:
\begin{itemize}
\item {\bf Constant kernel} A constant kernel component can be used to capture long-timescale
correlations in data, and can be represented with $\beta_R = \beta_I =  0$.
\item {\bf Exponential kernel.} This kernel, $K(\tau) = e^{-\vert \tau/\tau_0\vert}$, is exactly
represented with $\beta_I=0$, $\beta_R = \tau_0^{-1}$.
\item {\bf Cosine kernel.} The cosine kernel, $K(t) = \cos{\left(\frac{2\pi}{P} \vert \tau \vert\right)}$, 
may be represented exactly with $\beta_R=0$ and $\beta_I = \frac{2\pi}{P}$.
\item {\bf Squared exponential.} The squared exponential or Gaussian kernel, $G(z)=e^{-z^2/2}$,
has the behavior of declining steeply at large time separation, is an even function, and has $c_1$ = 0.
Nevertheless, it may be well approximated by the sum of three exponential kernels:
\begin{eqnarray}
G(z) &\approx& 2.751807 e^{-1.531250 z} \cos{(0.341489 z)} \\
 &-& 0.354626 e^{-4.753504 z} \cos{(2.042572 z)}\\
 &-& 1.396834 e^{-1.796743 z} \cos{(1.764168 z)},
\end{eqnarray}
which has an accuracy of better than $<4\times 10^{-4}$ for all z.
%[2.751807496353032,1.5312499408182438,-0.34148888180984543,-0.35462563655835033,4.753504292338051,2.0425716395901157,-1.3968344706423566,1.7967431129750422,-1.7641684073374964
\item {\bf Mat\'ern $p+1/2$ kernel.}  The Mat\'ern kernel is given by:
\begin{equation}
C_{p+1/2}(x) = \sigma^2 e^{-\sqrt{(2p+1)}x} \frac{\Gamma(p+1)}{\Gamma(2p+1)}
\sum_{i=0}^p \frac{(p+i)!}{i!(p-i)!} (\sqrt{(8p+4)}x)^{p-i},
\end{equation} 
where $p$ is a positive integer.  In the limit $p\rightarrow \infty$, this becomes the
squared exponential kernel: $C_\infty(x) = \sigma^2 e^{-x^2/2}$.  But, for small
values of $p$, the Mat\'ern kernel may be approximated well by the sum of
$2p$ exponential kernels.  The exponential kernel is the Mat\'ern 1/2 kernel, while
$p=1$ is relevant to stellar variability, which we describe next.
\item {\bf Mat\'ern 3/2 kernel.}  The Mat\'ern $3/2$ ($p=1$) kernel is given by: 
\begin{equation}
K_{3/2}(\tau) = C_{3/2}(\tau/\tau_0) = (1+v)e^{-v},
\end{equation} 
where $v = \sqrt{3}\vert \tau/\tau_0\vert$.  This kernel has a Fourier Transform\footnote{Using the normalization convention: 
$\hat f(\omega) = \frac{1}{\sqrt{2\pi}} \int_{-\infty}^\infty f(t) e^{i\omega t} dt$}  of
\begin{equation}
\hat K_{3/2}(\omega) = \frac{\tau_0 6^{3/2}}{\pi^{1/2}} (3+\tau_0^2\omega^2)^{-2},
\end{equation}
where $\omega = 2\pi f$.  This kernel may be well approximated by the sum of two exponential kernels: 
\begin{equation}
K_{3/2}(\tau) \approx (1-b)e^{-v}+be^{-v\frac{b-1}{b}},
\end{equation} 
where $b \gg 1$ is a dimensionless parameter; this approximation is exact in the limit $b \rightarrow \infty$.  For
large values of $b$, the maximum error of the approximation scales as $2/(e^2b) \approx 0.27/b$; for
$b=100$, the error in this approximation is $<0.3$\%.  Note that the two exponential terms
have opposite signs, which may lead to numerical instability for values of $b$ that
are large.

One advantage of the Mat\'ern 3/2 kernel is that it is smooth across
$\tau=0$ since its derivative is zero at $\tau=0$.  The exponential kernel {\it does not}
have this property, which leads to qualitatively different noise properties on small
timescales due to the sharpness of the exponential kernel at $\tau=0$.  The Mat\'ern 3/2 kernel
turns out to be similar to the observed properties of stellar activity and granulation, and thus the approximate
formula is simultaneously physically relevant and computationally convenient.  
The power spectrum of a single exponential scales as $f^{-2}$ at high frequency,
while the Mat\'ern kernel scales as $f^{-4}$ which more closely matches the power
spectrum due to activity and granulation in the Sun \citep{2004A&A...414.1139A,2009A&A...495..979M}.
\item {\bf Mat\'ern 5/2 kernel.}  The Mat\'ern 5/2 kernel is given by
\begin{equation}
K_{5/2}(t) = C_{5/2}(y) = \sigma^2 (1+y+y^2/3)e^{-y},
\end{equation}
where $y = \sqrt{5}\vert \tau/\tau_0\vert$.  The Fourier transform is:
\begin{equation}
\hat K_{5/2}(\omega) = \frac{2\tau_0 10^{5/2}}{3\pi^{1/2}} (5+\tau_0^2\omega^2)^{-3}.
\end{equation}
This kernel may be approximated well by the sum of four exponential kernels
\begin{eqnarray}
K_{5/2}(\tau) &\approx& -147.6965573 e^{-1.0097610 y} + 85.8875034 e^{-1.0621734 y}\\
 &-&4.6540017 e^{-0.8353745 y} + 67.4630744 e^{-0.9160394 y},
\end{eqnarray}
which is accurate to $<4\times 10^{-6}$ for $y \le 10$.
The Fourier transform has a steeper dependence at large frequencies, $f^{-6}$, and as we are not aware of a stellar
power spectrum that requires a steeper spectrum than this, we forego approximating further Mat\'ern kernels.
\item {\bf Exponential-sine-square kernel.}  This kernel is a periodic, positive valued kernel
which has been used to model quasi-periodic stellar variability due to star spots.   It is
given by $E(w,\Gamma) = e^{-\Gamma \sin^2{w}}$ where $w = \frac{\pi}{P}\vert \tau \vert$, which may be 
expanded as a series in $\cos{(2kw)} = \cos{\frac{2\pi}{P} k \vert \tau \vert)}$,
with $k \in \mathbb{N}$.  We have expanded this to $k=4$, and fit for the coefficients, giving:  
\begin{equation}
E(w,\Gamma) \approx \sum_{k=0}^4 \left(c_{k,1}+c_{k,2} e^{-c_{k,4}\Gamma^{c_{k,3}}}\right) \cos{(2kw)},
\end{equation}
with $c_{0,1}=0.24999, c_{0,2}=0.752803, c_{0,3}=0.960477, c_{0,4}=0.643393$ for $k=0$,
$c_{1,1}=0.450851, c_{1,2}=-0.445338, c_{1,3}=0.960477, c_{1,4}= 1.18102$ for $k=1$,
$c_{2,1}=0.206593, c_{2,2}=-0.207361, c_{2,3}=1.65902, c_{2,4}= 0.210641$ for $k=2$,
$c_{3,1}=0.0728449, c_{3,2}=-0.0729381, c_{3,3}=2.43149, c_{3,4}= 0.0471224$ for $k=3$, and
$c_{4,1}=0.019204, c_{4,2}=-0.0192129, c_{4,3}=3.27977, c_{4,4}= 0.0114198$ for $k=4$.
This approximation is accurate to better than 0.6\% for $\Gamma < 3$, with a standard
deviation of 0.1\%.
\end{itemize}

Examples of the foregoing kernels and their approximations are shown in Figure \ref{kernel_approx}.
It is clear from this figure that these standard kernels are well approximated by the sums
of between two and five damped cosine kernels, demonstrating that these form an adequate set
for modeling a wide range of kernel behaviors.

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=\textwidth]{kernel_approx.png}
\caption{Approximation of various kernels with the sums of damped cosines with $\tau_0=1$. 
(a). Kernels that may be expressed exactly: constant (dark grey), exponential (light grey), 
cosine (grey), and damped cosine (salmon).
(b). Approximation of the squared-exponential (Gaussian) kernel (pink) with three
damped cosine kernels (magenta). (c). The approximation of the exponential-sine-squared
kernel with $\Gamma =1$ (blue) and $\Gamma = 3$ (green). (d). Two Mat\'ern kernels,
$K_{3/2}$ (yellow/brown) and $K_{5/2}$ (violet).  The Mat\'ern 3/2 kernel
is approximated with $b=100$. }
\label{kernel_approx}
\end{center}
\end{figure}
